\section{Main Results}\label{sec:results}

\input{tables/overall_acc}

Table~\ref{tab:overall-acc} reports overall accuracy on 100 questions
for the full $4\times 8$ matrix of LLM families and backends. Three
patterns emerge.

\subsection{Production KG agents underperform in-context surface forms}

Across every LLM family, the in-context surface forms outperform the
production KG agents. Mean accuracy of the four in-context tiers
exceeds mean accuracy of the two KG agents by 8.25 percentage points
(Llama-4-Scout, the smallest gap), 22.0 (Qwen3-235B), 26.0
(gpt-oss-120b), and 28.75 (HCX-32B-Think, the largest).
Figure~\ref{fig:kg-vs-incontext} visualizes the gap.

\input{figures/main_bar}

This is counterintuitive in two ways. First, the KG agents
\emph{strictly} have access to more facts than the in-context dumps:
the production TypeDB and Neo4j stores contain entities that we did
not include in the in-context dumps (career tracks, allowance entities,
12 supplementary clauses, and three additional base-salary tables for
non-comprehensive-planning grades). Second, the KG agents are designed
exactly for multi-hop and multi-key queries — the regime in which the
in-context tiers should struggle most.

To test the gap rigorously we run paired McNemar exact two-sided
binomial tests on discordant pairs (Table~\ref{tab:mcnemar}). Of eight
KG-vs-context comparisons, five are statistically significant at
$p<0.05$ uncorrected, two are borderline ($p=0.052$), and only one
(Qwen-235B \backendname{typedb} vs \backendname{context}) fails to
reject the null. The win counts in the discordant column tell the
story: in HCX the KG agents win at most 5 questions while the context
agent wins 26.

\input{tables/mcnemar}

\paragraph{Multiple-testing correction.}
The picture is consistent under family-wise control. Bonferroni
correction at $\alpha/8 = 0.00625$ leaves three comparisons significant
(HCX \backendname{typedb}, HCX \backendname{neo4j}, Llama-4
\backendname{typedb}). Holm--Bonferroni concurs. Benjamini--Hochberg
FDR control at $\alpha=0.05$ yields six significant comparisons,
additionally Qwen \backendname{neo4j} and Llama-4 \backendname{neo4j}.
The headline finding — production KG agents under-performing
in-context tiers — survives the most conservative family-wise
correction in three of four LLM families.

\subsection{The query-generation stage is the bottleneck, not KG expressivity}

The KG agents and the in-context dumps differ in exactly one
operational respect: the KG agents \emph{generate a query} (TypeQL or
Cypher) whose results they then verbalize, while the in-context dumps
hand the full graph directly to the chat model. Since the in-context
dumps preserve the same hyper-relational role-player structure
(C-TQL) or property-graph structure (C-Cypher) that the KG agents
exploit operationally, any drop in KG-agent accuracy must be
attributed to the additional pipeline stages: query authoring, query
execution, and result re-verbalization. The 8--26 percentage-point
gap therefore quantifies the cumulative loss of those stages — the
\emph{query-generation bottleneck}.

\subsection{Closed-book leakage is low}

The \backendname{closed\_book} baselines achieve 9--13\% on the same
100 questions, with a four-LLM mean of 11.25\%. This bounds the
contribution of pre-training memorization to RAG performance: the
75--80 point gap between any RAG-style backend and
\backendname{closed\_book} is genuine retrieval, not regurgitation
of memorized facts.

\subsection{Natural-language summary Pareto-dominates structured surface forms}

Among the in-context tiers, C-NL achieves the highest mean accuracy
(89.8\% across LLMs) at less than half the token cost of the
structured tiers (14K characters vs.\ 26--32K).
Table~\ref{tab:token-cost} and Figure~\ref{fig:pareto} display the
trade-off.

\input{tables/token_cost}
\input{figures/nl_pareto}

The structured tiers (TQL, Cypher, JSON-LD) bring no accuracy benefit
proportional to their token overhead. The single exception —
gpt-oss-120b on C-TQL at 95\% — does not generalize: Llama-4-Scout
falls to 63\% on the same surface form, dragging down the C-TQL
mean. C-NL is the only surface form that maintains $\geq 83\%$
accuracy across all four LLMs.

\subsection{On Setting A vs.\ Setting B}

A 30-question pilot under \emph{Setting A} — same LLM as chat and
query agent — produced LLM-family-asymmetric KG results: Qwen3-235B
on \backendname{neo4j} jumped +23\,pp from Setting A to Setting B,
while \backendname{typedb} dropped --6\,pp. The asymmetry confirms
that single-LLM measurements conflate KG-backend effects with the
LLM's TypeQL/Cypher syntax familiarity, which is precisely the
mechanism Section~\ref{sec:mechanism} dissects. We therefore use
Setting B (query agent fixed at Qwen3-Coder-30B) as the
production-realistic configuration throughout this paper. Full
Setting A figures appear in Appendix~\ref{sec:appendix}; we report
them transparently rather than as a contrast that flatters the
main result.

\subsection{Multi-key categories: the regime KG agents are designed for}

Restricting to the H1-relevant categories — \texttt{calculation},
\texttt{comparison}, \texttt{multihop\_3key} (42 questions total) —
sharpens the picture: the KG agents range from 47.6\% to 66.7\%
on this subset (mean 57\%) while the in-context tiers reach
57.1\%--100\% (mean 90\%). The regime that hyper-relational KGs are
operationally tailored for is precisely where the in-context
full-graph dumps win largest, with per-LLM multi-key gaps of
17.9, 32.7, 39.9, and 41.7 percentage points (Llama-4-Scout,
Qwen3-235B, HCX-32B-Think, gpt-oss-120b respectively).
