\section{Conclusion}\label{sec:conclusion}

We have shown, on a 100-question Korean public-sector regulation
benchmark with four LLM families, that the production KG-RAG
agents — built on TypeDB v3 and Neo4j — underperform an in-context
surface-form ablation of the \emph{same} knowledge graph by
8--26 percentage points (paired McNemar $p<0.05$ in 5 of 8
KG-vs-context comparisons). The result holds despite the agents
having strictly more data and being deployed for exactly the
multi-key reasoning regime they appear designed to dominate.

The bottleneck is the query-generation stage of the KG agent, not
the expressivity of the KG itself: an in-context dump that
preserves the same hyper-relational role-player structure reaches
83--95\% accuracy. Within that in-context regime, natural-language
summary Pareto-dominates the formal surface forms (TypeQL, Cypher,
JSON-LD) on both accuracy and token cost, and the only LLM family
sensitive to structural surface-form (Llama-4-Scout) reveals a
secondary mechanism: pre-training query-language familiarity that
syntax guides only partly compensate for. A third mechanism — the
\emph{KG semantic paradox} — shows that the very structural clarity
that makes a KG operationally valuable disappears when serialized,
hurting closed-world boundary inference relative to natural
language.

We argue that KG infrastructure should be treated as the
\emph{operational backbone} of the system — schema enforcement,
multi-system integration, query-agent improvement potential — and
that its LLM-facing layer should be a structure-preserving
natural-language summary, augmented with explicit boundary
annotations. The reframing is constructive rather than dismissive:
KGs remain irreplaceable for the operational dimensions, and the
gap between the production KG agent and the in-context tier is a
roadmap for stronger query-generation models, not a verdict on
KG expressivity.

\paragraph{Future work.}
We outline three priorities in Section~\ref{sec:limits}:
multi-domain replication beyond Korean public-sector regulation,
explicit-boundary NL variants with quantified lift on the
out-of-scope category, and external-baseline comparison
(GraphRAG, HippoRAG) on the same evaluation set.

\paragraph{Reproducibility.}
The evaluation set, scorer, runner, and surface-form dump
generators are released at
\url{https://github.com/shezgone/bok-comp-eval}. Three of the four
LLM families used (Qwen3-235B, Llama-4-Scout, gpt-oss-120b) are
open-weight; replication on their HuggingFace checkpoints
reproduces the in-context-tier numbers.
